Follow-up Samples

### 2016-08-02 Baxter_followUps_2016.Rmd
removed unused chunk

---
title: "Follow Up Samples"
author: "Niel"
date: "November 20, 2015"
output: 
  html_document:
    keep_md: true
---

@@

```{r dists, echo=FALSE, fig.height=5, fig.width=10}
source('code/read.dist.R')

@@

set.seed(112015)
rf <- AUCRF(lesion~., data=train, pdel=0.1, ntree=1000, ranking='MDA')
rf_opt <- rf$RFopt
#plot(rf)

@@

```{r fit_dif, echo=FALSE, fig.height=6, fig.width=6, warning=FALSE}
plot(0, type='n', xlim=c(0.5,4.5), ylim=c(0.05,5000), log='y', ylab='FIT Results (ng/ml)', xaxt='n', xlab='', yaxt='n')
mtext(text=c('Adenoma','Cancer'), at=c(1.5,3.5), side=1, line=2, font=2)
axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow Up','Baseline','Follow Up'))
axis(2, at=c(0.1,1.1,10.1,100.1,1000.1), labels=c(0,1,10,100,1000))

@@

```{r comb_dif, echo=FALSE, fig.height=6, fig.width=6}
plot(0, type='n', xlim=c(0.5,4.5), ylim=c(0,1.1), ylab='Probability of lesion (Combined model)', xaxt='n', xlab='')
mtext(text=c('Adenoma','Cancer'), at=c(1.5,3.5), side=1, line=2, font=2)
axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow Up','Baseline','Follow Up'))

points(rep(3,26), initial_probs[metaF$dx=='cancer'], pch=21, bg='red', cex=0.8, lwd=0.5)
points(rep(4,26), follow_probs[metaF$dx=='cancer'], pch=21, bg='red', cex=0.8, lwd=0.5)
segments(3, initial_probs[metaF$dx=='cancer'], 4, follow_probs[metaF$dx=='cancer'], col='red')
canc_p <- t.test(initial_probs[metaF$dx=='cancer'], follow_probs[metaF$dx=='cancer'], paired=T, alternative='greater')$p.value
text(3.5, 1, sprintf('p=%.1g',canc_p))

@@

```{r otu_model, echo=FALSE, cache=TRUE, fig.height=5, fig.width=6}
# Make training set with samples withouth follow ups
train <- merge(metaI, shared, by.x='sample', by.y='Group')
train <- train[-match(metaF$initial, train$sample),]
train <- train[,c('lesion',colnames(train)[grep('Otu[0123456789]', colnames(train))])]

@@

set.seed(112015)
rf_otu <- AUCRF(lesion~., data=train, pdel=0.1, ntree=1000, ranking='MDA')
rf_opt <- rf_otu$RFopt
#plot(rf_otu)

@@

```{r dif_otu, echo=FALSE, fig.height=6, fig.width=6}
plot(0, type='n', xlim=c(0.5,4.5), ylim=c(0,1.1), ylab='Probability of lesion (OTU model)', xaxt='n', xlab='')
mtext(text=c('Adenoma','Cancer'), at=c(1.5,3.5), side=1, line=2, font=2)
axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow Up','Baseline','Follow Up'))

@@

points(rep(3,26), initial_probs[metaF$dx=='cancer'], pch=21, bg='red', cex=0.8, lwd=0.5)
points(rep(4,26), follow_probs[metaF$dx=='cancer'], pch=21, bg='red', cex=0.8, lwd=0.5)
segments(3, initial_probs[metaF$dx=='cancer'], 4, follow_probs[metaF$dx=='cancer'], col='red')
canc_p <- t.test(initial_probs[metaF$dx=='cancer'], follow_probs[metaF$dx=='cancer'], paired=T, alternative='greater')$p.value
text(3.5, 1, sprintf('p=%.1g',canc_p))

@@

@@

set.seed(112015)
rf_otu <- AUCRF(type~., data=adeVcanc, pdel=0.1, ntree=1000, ranking='MDA')
rf_opt <- rf_otu$RFopt

@@

```{r otu_abunds, echo=FALSE, fig.height=6, fig.width=6}
i_abunds <- initial[,-c(1,2)]/10530 + 0.0001
f_abunds <- followups[,-c(1,2)]/10530 + 0.0001
taxa <- c('P. asaccharolytica','P. stomatis','F. nucleatum', 'P. micra')

@@

pdf('figures/otu_abunds.pdf', height=4, width=8)
layout(matrix(c(1:4), nrow=1))
par(mar=c(4,0.5,1,0.5), oma=c(0,4,0,0))
for(i in 1:length(otus)){
  plot(0, type='n', log='y', xlim=c(0.75,2.25), ylim=c(0.0001,0.2), ylab='', xaxt='n', xlab='', yaxt='n', las=2)
  points(rep(1,nrow(i_abunds)), i_abunds[,otus[i]], pch=21, bg=c('orange','red')[metaF$dx], cex=1, lwd=0.5)
  points(rep(2,nrow(f_abunds)), f_abunds[,otus[i]], pch=21, bg=c('orange','red')[metaF$dx], cex=1, lwd=0.5)
  segments(1, i_abunds[,otus[i]], 2, f_abunds[,otus[i]], col=c('orange','red')[metaF$dx])
  axis(1, at=c(1,2), c('Initial','Follow Up'))
  mtext(taxa[i], 1, line=2.5, cex=0.8)
  p_c<-wilcox.test(initial[metaF$dx=='cancer',otus[i]], followups[metaF$dx=='cancer',otus[i]], paired=T, alternative='greater')$p.value
  text(1.5,0.19, labels=sprintf('Carcinomas: p=%.2g', p_c))
  p_a<-wilcox.test(initial[metaF$dx=='adenoma',otus[i]], followups[metaF$dx=='adenoma',otus[i]], paired=T, alternative='greater')$p.value
  text(1.5,0.13, labels=sprintf('Adenomas: p=%.2g', p_a))
  if(i==1){axis(2, at=c(0.0001,0.001,0.01,0.1,1), c(0,0.1,1,10,100))
      mtext('% Relative Abundace', 2, line=2.5, cex=0.8)}
  #if(i==4){legend('top', c('Adenoma','Carcinoma'), pch=21, pt.bg=c('orange','red'))}
}
dev.off()
```

>>>>

---
title: "Follow-up Samples"
author: "Niel"
date: "November 20, 2015"
output: 
  html_document:
    keep_md: true
---

@@

```{r dists, echo=FALSE, fig.height=5, fig.width=10, cache=T}
source('code/read.dist.R')

@@

set.seed(050416)
rf <- AUCRF(lesion~., data=train, pdel=0.05, ntree=500, ranking='MDA')
rf_opt <- rf$RFopt
#plot(rf)

@@

```{r fit_dif, echo=FALSE, fig.height=6, fig.width=6, warning=FALSE}
plot(1, type='n', xlim=c(0.5,4.5), ylim=c(0.05,5000), log='y', ylab='FIT Results (ng/ml)', xaxt='n', xlab='', yaxt='n')
mtext(text=c('Adenoma','Cancer'), at=c(1.5,3.5), side=1, line=2, font=2)
axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow-up','Baseline','Follow-up'))
axis(2, at=c(0.1,1.1,10.1,100.1,1000.1), labels=c(0,1,10,100,1000))

@@

```{r comb_dif, echo=FALSE, fig.height=6, fig.width=6}
pdf('figures/mmt_dif.pdf', height=6, width=8)
layout(1)
par(mgp=c(2.5,0.5,0))
plot(0, type='n', xlim=c(0.5,4.5), ylim=c(0,1), ylab='Probability of lesion (Combined model)', xaxt='n', xlab='', las=2)
mtext(text=c('Adenoma','Cancer'), at=c(1.5,3.5), side=1, line=2, font=2)
axis(1, at=c(1,2,3,4), labels=c('Initial','Follow-up','Initial','Follow-up'))

jit <- jitter(rep(0,41), amount=0.05)
points(rep(1,41)+jit, initial_probs[metaF$dx=='adenoma'], pch=21, bg='orange', cex=1, lwd=0.5)
points(rep(2,41)+jit, follow_probs[metaF$dx=='adenoma'], pch=21, bg='orange', cex=1, lwd=0.5)
segments(1, initial_probs[metaF$dx=='adenoma'], 2, follow_probs[metaF$dx=='adenoma'], col='orange')
ade_p<-t.test(initial_probs[metaF$dx=='adenoma'], follow_probs[metaF$dx=='adenoma'], paired=T, alternative='greater')$p.value
text(1.5, 1, sprintf('p=%.1g',ade_p))

jit <- jitter(rep(0,26), amount=0.05)
points(rep(3,26)+jit, initial_probs[metaF$dx=='cancer'], pch=21, bg='red', cex=1, lwd=0.5)
points(rep(4,26)+jit, follow_probs[metaF$dx=='cancer'], pch=21, bg='red', cex=1, lwd=0.5)
segments(3, initial_probs[metaF$dx=='cancer'], 4, follow_probs[metaF$dx=='cancer'], col='red')
canc_p <- t.test(initial_probs[metaF$dx=='cancer'], follow_probs[metaF$dx=='cancer'], paired=T, alternative='greater')$p.value
text(3.5, 1, sprintf('p=%.1g',canc_p))

abline(h=cutoff, lty=2)
dev.off()
```

@@

```{r otu_model, echo=FALSE, cache=TRUE, fig.height=5, fig.width=6, eval=T}
# Make training set with samples withouth follow ups
train <- merge(metaI, shared, by.x='sample', by.y='Group')
train <- train[-which(train$sample %in% metaF$initial),]
train <- train[,c('lesion',colnames(train)[grep('Otu[0123456789]', colnames(train))])]

@@

set.seed(112015)
rf_otu <- AUCRF(lesion~., data=train, pdel=0.1, ntree=500, ranking='MDA')
rf_opt <- rf_otu$RFopt

@@

```{r dif_otu, echo=FALSE, fig.height=6, fig.width=6}

plot(0, type='n', xlim=c(0.5,4.5), ylim=c(0,1.1), ylab='Probability of lesion (OTU model)', xaxt='n', xlab='')
mtext(text=c('Adenoma','Cancer'), at=c(1.5,3.5), side=1, line=2, font=2)
axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow-up','Baseline','Follow-up'))

@@

points(rep(3,26), initial_probs[metaF$dx=='cancer'], pch=c(21,22,23,24,25)[metaF$stage], bg='red', cex=0.8, lwd=0.5)
points(rep(4,26), follow_probs[metaF$dx=='cancer'], pch=c(21,22,23,24,25)[metaF$stage], bg='red', cex=0.8, lwd=0.5)
segments(3, initial_probs[metaF$dx=='cancer'], 4, follow_probs[metaF$dx=='cancer'], col='red')
canc_p <- t.test(initial_probs[metaF$dx=='cancer'], follow_probs[metaF$dx=='cancer'], paired=T, alternative='greater')$p.value
text(3.5, 1, sprintf('p=%.1g',canc_p))

@@

```{r diff_by_stage, echo=FALSE, cache=TRUE, fig.height=5, fig.width=6}
# Make training set with samples without follow ups
train <- merge(metaI, shared, by.x='sample', by.y='Group')
train <- train[-which(train$sample %in% metaF$initial),]
train <- train[, c('lesion',colnames(train)[grep('Otu[0123456789]', colnames(train))])]

set.seed(050216)
rf_otu <- AUCRF(lesion~., data=train, pdel=0.1, ntree=500, ranking='MDA')
rf_opt <- rf_otu$RFopt
#plot(rf_otu)

layout(matrix(c(1,2,3,4), nrow=1))
par(mar=c(5,3,1,0.5))
#non advanced
plot(0, type='n', xlim=c(0.75,2.25), ylim=c(0,1.1), ylab='Probability of lesion (OTU model)', xaxt='n', xlab='')
mtext(text='Nonadvanced\nAdenoma', at=c(1.5), side=1, line=4, font=2)
axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow-up','Baseline','Follow-up'))
stripchart(at=1, initial_probs[metaF$Dx=='Adenoma'], pch=21, bg='orange', cex=1.2, lwd=0.5, vertical=T, add=T)
stripchart(at=2, follow_probs[metaF$Dx=='Adenoma'], pch=21, bg='orange', cex=1.2, lwd=0.5, vertical=T, add=T)
segments(1, initial_probs[metaF$Dx=='Adenoma'], 2, follow_probs[metaF$Dx=='Adenoma'], col='orange')
ade_p<-t.test(initial_probs[metaF$Dx=='Adenoma'], follow_probs[metaF$Dx=='Adenoma'], paired=T, alternative='greater')$p.value
text(1.5, 1, sprintf('p=%.1g',ade_p))

#advanced
plot(0, type='n', xlim=c(0.75,2.25), ylim=c(0,1.1), ylab='Probability of lesion (OTU model)', xaxt='n', xlab='')
mtext(text='Advanced\nAdenoma', at=c(1.5), side=1, line=4, font=2)
axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow-up','Baseline','Follow-up'))
stripchart(at=1, initial_probs[metaF$Dx=='adv Adenoma'], pch=21, bg='orange', cex=1.2, lwd=0.5, vertical=T, add=T)
stripchart(at=2, follow_probs[metaF$Dx=='adv Adenoma'], pch=21, bg='orange', cex=1.2, lwd=0.5, vertical=T, add=T)
segments(1, initial_probs[metaF$Dx=='adv Adenoma'], 2, follow_probs[metaF$Dx=='adv Adenoma'], col='orange')
ade_p<-t.test(initial_probs[metaF$Dx=='adv Adenoma'], follow_probs[metaF$Dx=='adv Adenoma'], paired=T, alternative='greater')$p.value
text(1.5, 1, sprintf('p=%.1g',ade_p))

#Early (stage 1 or 2)
plot(0, type='n', xlim=c(0.75,2.25), ylim=c(0,1.1), ylab='Probability of lesion (OTU model)', xaxt='n', xlab='')
mtext(text='Cancer\nStage I or II', at=c(1.5), side=1, line=4, font=2)
axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow-up','Baseline','Follow-up'))
stripchart(at=1, initial_probs[metaF$stage==1 | metaF$stage==2], pch=21, bg='red', cex=1.2, lwd=0.5, vertical=T, add=T)
stripchart(at=2, follow_probs[metaF$stage==1 | metaF$stage==2], pch=21, bg='red', cex=1.2, lwd=0.5, vertical=T, add=T)
segments(1, initial_probs[metaF$stage==1 | metaF$stage==2], 2, follow_probs[metaF$stage==1 | metaF$stage==2], col='red')
ade_p<-t.test(initial_probs[metaF$stage==1 | metaF$stage==2], follow_probs[metaF$stage==1 | metaF$stage==2], paired=T, alternative='greater')$p.value
text(1.5, 1, sprintf('p=%.1g',ade_p))

#Early (stage 3 or 4)
plot(0, type='n', xlim=c(0.75,2.25), ylim=c(0,1.1), ylab='Probability of lesion (OTU model)', xaxt='n', xlab='')
mtext(text='Cancer\nStage III or IV', at=c(1.5), side=1, line=4, font=2)
axis(1, at=c(1,2,3,4), labels=c('Baseline','Follow-up','Baseline','Follow-up'))
stripchart(at=1, initial_probs[metaF$stage==3 | metaF$stage==4], pch=21, bg='red', cex=1.2, lwd=0.5, vertical=T, add=T)
stripchart(at=2, follow_probs[metaF$stage==3 | metaF$stage==4], pch=21, bg='red', cex=1.2, lwd=0.5, vertical=T, add=T)
segments(1, initial_probs[metaF$stage==3 | metaF$stage==4], 2, follow_probs[metaF$stage==3 | metaF$stage==4], col='red')
ade_p<-t.test(initial_probs[metaF$stage==3 | metaF$stage==4], follow_probs[metaF$stage==3 | metaF$stage==4], paired=T, alternative='greater')$p.value
text(1.5, 1, sprintf('p=%.1g',ade_p))

@@

set.seed(112015)
rf_otu <- AUCRF(type~., data=adeVcanc, pdel=0.1, ntree=500, ranking='MDA')
rf_opt <- rf_otu$RFopt

@@

```{r otu_abunds, echo=FALSE, message=F, warning=F, fig.height=4, fig.width=7}
i_abunds <- initial[,-c(1,2)]/10530 + 0.0001
f_abunds <- followups[,-c(1,2)]/10530 + 0.0001
taxa <- c('P. asaccharolytica','P. stomatis','F. nucleatum', 'P. micra')

@@

layout(matrix(c(1:4), nrow=1))
par(mar=c(4,0.5,1,0.5), oma=c(0,4,0,0))

jity<-jitter(rep(0,67), amount=0.00001)
jitx<-jitter(rep(0,67), amount=0.05)

for(i in 1:length(otus)){
  plot(0, type='n', log='y', xlim=c(0.75,2.25), ylim=c(0.00009,0.2), ylab='', xaxt='n', xlab='', yaxt='n', las=2)
  points(rep(1,nrow(i_abunds))+jitx, jity+i_abunds[,otus[i]], pch=21, bg=c('orange','red')[metaF$dx], cex=1, lwd=0.5)
  points(rep(2,nrow(f_abunds))+jitx, jity+f_abunds[,otus[i]], pch=21, bg=c('orange','red')[metaF$dx], cex=1, lwd=0.5)
  segments(1+jitx, i_abunds[,otus[i]], 2+jitx, f_abunds[,otus[i]], col=c('orange','red')[metaF$dx])
  axis(1, at=c(1,2), c('Initial','Follow-up'))
  mtext(taxa[i], 1, line=2.5, cex=0.8)
  p_c<-wilcox.test(initial[metaF$dx=='cancer',otus[i]], followups[metaF$dx=='cancer',otus[i]], paired=T, alternative='greater')$p.value
  text(1.5,0.19, labels=sprintf('Carcinomas: p=%.2g', p_c))
  p_a<-wilcox.test(initial[metaF$dx=='adenoma',otus[i]], followups[metaF$dx=='adenoma',otus[i]], paired=T, alternative='greater')$p.value
  text(1.5,0.13, labels=sprintf('Adenomas: p=%.2g', p_a))
  if(i==1){axis(2, at=c(0.0001,0.001,0.01,0.1,1), c(0,0.1,1,10,100))
      mtext('% Relative Abundace', 2, line=2.5, cex=0.8)}
  #if(i==4){legend('top', c('Adenoma','Carcinoma'), pch=21, pt.bg=c('orange','red'))}
}
```

**Figure 5.** I compared the abundance of the four main cancer-associated OTUs over time. There was a significant decrease in the abundance of these OTUs for patients with carcinomas, but for patients with adenomas.
